\chapter{LLaMA-Class Architectures}
\label{ch:llama-class-architectures}

\abstract{This chapter studies the architectural conventions used by modern open decoder models: RMSNorm, RoPE, SwiGLU-style feed-forward networks, grouped-query attention, KV cache design, long-context extensions, and mixture-of-experts variants. LLaMA-class models are not a different species from GPT; they are a refined decoder-only recipe whose details change stability, memory, throughput, and context behavior.}

\section{What Makes a Decoder LLaMA-Class}
\label{sec:what-makes-llama-class}

Modern open \llms are mostly decoder-only \transformers, but the defaults have changed since early GPT implementations. LLaMA and its successors popularized a compact recipe: pre-normalized decoder blocks, RMSNorm, rotary position embeddings, bias-free linear layers, SwiGLU-style feed-forward networks, careful tokenizer and data choices, and post-training pipelines that make the base model usable as an assistant \cite{touvron2023llama,touvron2023llama2,dubey2024llama3}. Later families add grouped-query attention, longer context, stronger multilingual and code mixtures, tool-oriented post-training, and sparse mixture-of-experts variants \cite{ainslie2023gqa,deepseek2024v3,kimi2025k2}.

A LLaMA-class block can be written as
\begin{align}
  y^{(\ell)}
    &= x^{(\ell)} +
       \mathrm{Attn}\left(\mathrm{RMSNorm}(x^{(\ell)})\right),\\
  x^{(\ell+1)}
    &= y^{(\ell)} +
       \mathrm{SwiGLU}\left(\mathrm{RMSNorm}(y^{(\ell)})\right).
\end{align}
This is still the residual-stream structure from Chapter~\ref{ch:transformer-mechanics}. The significance is in the defaults. RMSNorm changes normalization cost and behavior; RoPE changes where positions enter attention; SwiGLU changes MLP expressivity and parameter allocation; grouped-query attention changes the KV-cache budget. These are not cosmetic differences when training or serving billions of parameters.

\begin{figure}[t]
\centering
\begin{tikzpicture}[node distance=6mm and 7mm]
\node[llmblock, text width=2.2cm] (x) {Residual\\stream};
\node[llmblock, right=of x, text width=2.0cm] (r1) {RMSNorm};
\node[llmblock, right=of r1, text width=2.2cm] (rope) {RoPE on\\$Q,K$};
\node[llmblock, right=of rope, text width=2.2cm] (gqa) {GQA\\attention};
\node[llmblock, below=of gqa, text width=2.2cm] (r2) {RMSNorm};
\node[llmblock, left=of r2, text width=2.2cm] (ffn) {SwiGLU\\or MoE FFN};
\node[llmblock, left=of ffn, text width=2.2cm] (out) {Next residual\\stream};
\draw[llmarrow] (x) -- (r1);
\draw[llmarrow] (r1) -- (rope);
\draw[llmarrow] (rope) -- (gqa);
\draw[llmarrow] (gqa) -- (r2);
\draw[llmarrow] (r2) -- (ffn);
\draw[llmarrow] (ffn) -- (out);
\draw[llmdash] (x.south) .. controls +(0,-1.0) and +(0,-1.0) .. node[below, font=\scriptsize] {residual} (gqa.south);
\draw[llmdash] (gqa.west) .. controls +(-0.8,-1.0) and +(0.8,-1.0) .. node[below, font=\scriptsize] {residual} (out.east);
\node[llmnote, below=8mm of ffn, text width=0.78\textwidth] {The LLaMA-class recipe is a set of coupled defaults. Changing GQA changes serving memory; changing RoPE changes long-context behavior; changing SwiGLU or MoE changes parameter allocation and communication.};
\end{tikzpicture}
\caption{A LLaMA-class decoder block as a contract among normalization, position handling, attention cache structure, and feed-forward capacity. The drawing abstracts from LLaMA, Llama 2/3, DeepSeek-V3, and Kimi K2-style reports without copying any released architecture figure.}
\label{fig:llama-class-block}
\end{figure}

\section{RMSNorm and Pre-Normalization}
\label{sec:llama-rmsnorm-prenorm}

RMSNorm rescales a hidden vector without subtracting its mean \cite{zhang2019root}. For $h\in\mathbb{R}^{d}$,
\begin{equation}
  \mathrm{RMSNorm}(h)
  = g \odot
    \frac{h}{\sqrt{\frac{1}{d}\sum_{j=1}^{d}h_j^2+\epsilon}},
  \label{eq:rmsnorm-llama}
\end{equation}
where $g$ is a learned scale vector. Compared with LayerNorm, RMSNorm removes the mean reduction and bias term in common implementations. The saved work is small per vector but large across every token, layer, and generation request. More importantly, it is part of a block recipe that has been empirically stable for deep decoder-only models.

Pre-normalization means the attention and MLP sublayers see normalized inputs, while residual additions carry unnormalized accumulated state forward. This tends to improve optimization at depth because each sublayer receives a controlled scale even as the residual stream evolves. The final RMSNorm before the language-model head is also important: it makes logits depend on a normalized representation rather than on raw residual scale.

\paragraph{Implementation note.}
RMSNorm is easy to implement incorrectly in mixed precision. The mean of squares should usually be accumulated in a sufficiently stable dtype, and $\epsilon$ should match the checkpoint family. Changing $\epsilon$ or whether the scale is applied before or after casting can produce small logit differences that become visible during long autoregressive generation.

\section{SwiGLU Feed-Forward Networks}
\label{sec:llama-swiglu}

The simple GPT MLP expands from $d$ to $4d$, applies GELU, and projects back to $d$. LLaMA-class models often use a gated feed-forward network inspired by GLU variants \cite{shazeer2020glu}:
\begin{equation}
  \mathrm{FFN}(h)
  = W_o\left(\mathrm{SiLU}(W_g h) \odot W_u h\right).
  \label{eq:swiglu-llama}
\end{equation}
There are two input projections, $W_g$ and $W_u$, and one output projection, $W_o$. To keep parameter count near the ungated $4d$ MLP, the hidden width is often set near $\frac{8}{3}d$ and rounded to a hardware-friendly multiple. The gate lets the model modulate candidate features elementwise, while the up projection supplies the candidate values.

The feed-forward block is often the largest parameter component of a dense decoder layer. With hidden width $m$, the gated MLP has approximately $3dm$ weights, ignoring biases. A standard $4d$ MLP has approximately $8d^2$ weights. Setting $m\approx \frac{8}{3}d$ makes these comparable. Rounding $m$ upward can improve tensor-core utilization but changes parameter count, activation memory, and checkpoint size.

\section{Rotary Position Embeddings}
\label{sec:llama-rope}

RoPE applies a position-dependent rotation to query and key vectors before the attention dot product \cite{su2021roformer}. Pair adjacent dimensions and rotate each pair by angle $m\theta_i$ at position $m$:
\begin{equation}
  R_{m,i} =
  \begin{bmatrix}
  \cos(m\theta_i) & -\sin(m\theta_i)\\
  \sin(m\theta_i) & \cos(m\theta_i)
  \end{bmatrix}.
\end{equation}
If a query at position $m$ and a key at position $n$ are rotated, their dot product contains
\begin{equation}
  (R_m q)^{\top}(R_n k)
  = q^{\top}R_m^{\top}R_n k,
\end{equation}
and $R_m^{\top}R_n$ depends on the relative offset $n-m$. This is the geometric reason RoPE gives attention access to relative position while remaining compatible with standard dot-product attention.

RoPE is not a free long-context solution. A model trained up to one context length has learned to use a particular range of rotation phases. Extrapolating to much longer contexts can create phase patterns that were rare or absent during training. Long-context extensions use methods such as position interpolation, frequency scaling, continued training, sliding attention, or attention biases. ALiBi is a contrasting approach that adds linear distance biases rather than rotating query and key vectors \cite{press2021train}. The evaluation question is not whether a model accepts a longer input tensor, but whether it uses distant evidence reliably without damaging shorter-context behavior.

\paragraph{Implementation note.}
RoPE caches must align with the decode cache. In prefill, positions are usually $0,\ldots,T-1$. In cached decoding, the new token's query and key must use the next logical position, not position 0. A one-token error can be almost invisible on short prompts and severe on long generations.

\section{KV Cache and Attention Variants}
\label{sec:llama-kv-cache-attention-variants}

During autoregressive decoding, every layer stores keys and values from previous tokens. For a dense model with $L$ layers, batch $B$, context length $T$, head dimension $d_h$, and $H_{\mathrm{kv}}$ key-value heads, the approximate cache memory is
\begin{equation}
  \mathrm{bytes}_{\mathrm{KV}}
  = 2BLT H_{\mathrm{kv}} d_h s,
  \label{eq:llama-kv-cache}
\end{equation}
where $s$ is bytes per element. Multi-head attention has $H_{\mathrm{kv}}=H_q$, one KV head per query head. Multi-query attention shares a single KV head across all query heads, reducing cache memory and memory bandwidth during decoding \cite{shazeer2019fast}. Grouped-query attention uses an intermediate number of KV heads, grouping several query heads per KV head, and often preserves more quality than full multi-query attention at similar serving benefits \cite{ainslie2023gqa}.

\begin{table}[t]
\caption{Attention variants by KV-cache structure.}
\label{tab:attention-variants-kv}
\small
\begin{tabular}{@{}p{2.8cm}p{2.2cm}p{3.5cm}p{3.1cm}@{}}
\hline\noalign{\smallskip}
Variant & KV Heads & Main Benefit & Main Risk \\
\noalign{\smallskip}\svhline\noalign{\smallskip}
Multi-head attention & $H_q$ & Maximum per-head KV flexibility & Largest decode cache and bandwidth \\
Multi-query attention & $1$ & Very small KV cache & Quality loss on some tasks or scales \\
Grouped-query attention & $1 < H_{\mathrm{kv}} < H_q$ & Good memory-quality tradeoff & Requires compatible checkpoint design \\
Latent or compressed attention & Model-specific & Smaller or transformed memory state & Harder to compare semantics and kernels \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

Equation~\ref{eq:llama-kv-cache} explains why grouped-query attention became a practical default. Weight memory is paid once per loaded model replica, but KV memory grows with active batch and context. In a long-context serving system, KV cache can dominate capacity. Reducing $H_{\mathrm{kv}}$ improves throughput not only by saving memory, but also by reducing memory bandwidth in token-by-token decoding.

Some newer architectures go beyond GQA with latent or compressed attention states, specialized cache layouts, or sparse attention patterns \cite{deepseek2024v3}. These variants should be evaluated under the exact serving regime of interest. A method that improves peak throughput for long prompts may add overhead for short prompts; a cache compression that preserves average benchmark scores may fail on retrieval-heavy tasks requiring exact distant evidence.

\section{Tokenizer and Vocabulary Contracts}
\label{sec:llama-tokenizer-vocabulary-contracts}

LLaMA-class checkpoints inherit the tokenizer-model coupling from Chapter~\ref{ch:tokens-corpora-training-signals}. The vocabulary determines embedding and output projection shapes, special-token ids, chat-template conventions, and token efficiency. Open model families often change tokenizer vocabularies across generations to improve multilingual and code coverage. Such changes are architectural from the checkpoint's point of view: the embedding table, output head, and all learned token statistics are tied to the tokenizer.

Vocabulary size is also a systems parameter. Embedding and output matrices scale as $Vd_{\mathrm{model}}$. During training, the final softmax over $V$ can be expensive; during inference, logits processors and sampling operate over $V$ each step. Implementations often pad vocabulary size to a multiple that improves matrix-kernel efficiency. Padding rows must not become ordinary sampled tokens.

\section{Long Context}
\label{sec:llama-long-context}

Long context has three separate meanings. The first is acceptance: the model can run a forward pass at length $T$. The second is retrieval: the model can identify relevant information somewhere in that context. The third is reasoning or synthesis: the model can combine distant evidence correctly. Many failures occur because a system demonstrates the first and claims the third.

Training for long context increases attention cost, activation memory, data requirements, and evaluation burden. Continuing training at longer sequence lengths can adapt positional behavior, but it may also change short-context performance. Synthetic needle-in-a-haystack tests are useful sanity checks but incomplete. A model can find a unique key phrase and still fail when evidence is paraphrased, conflicting, tabular, or distributed across sections. Long-context evaluation should include retrieval, aggregation, contradiction handling, citation faithfulness, and latency/cost reporting.

Serving long context is dominated by prefill and cache pressure. Prefill cost grows with the prompt length and full attention pattern; decode cost grows with cache length. Sliding-window attention, retrieval-augmented generation, summary memory, and cache eviction policies are alternatives to simply increasing $T_{\max}$. These alternatives change the task semantics and should be exposed in evaluation.

\section{Mixture-of-Experts Variants}
\label{sec:llama-moe-variants}

Mixture-of-experts (MoE) models replace some dense feed-forward computation with a set of expert networks and a router. For token hidden state $h_t$, a router produces scores over experts. A top-$k$ router selects a small subset $S_t$ and computes
\begin{equation}
  \mathrm{MoE}(h_t)
  = \sum_{e\in S_t} \alpha_{t,e} E_e(h_t),
\end{equation}
where $E_e$ is expert $e$ and $\alpha_{t,e}$ is a routing weight. The model can have many total parameters while activating only a fraction per token. This changes the meaning of model size: total parameters, active parameters, routing overhead, and communication cost must be reported separately.

Sparse computation has failure modes that dense models do not. Routers can collapse onto a few experts, leaving others undertrained. Load-balancing losses can improve hardware utilization while interfering with specialization. Expert parallelism introduces all-to-all communication across devices. Capacity limits can drop or reroute tokens. Evaluation can hide these problems if it reports only aggregate accuracy rather than per-domain behavior, routing distribution, latency percentiles, and failure cases.

MoE models such as DeepSeek-V3 and Kimi K2-style systems show how sparse scaling can support strong open-weight performance claims, but the systems contract is more complex than a dense LLaMA-class model \cite{deepseek2024v3,kimi2025k2}. A publishable comparison should specify active parameters, total parameters, context length, data mixture where known, post-training recipe, inference budget, and serving hardware.

\section{Beyond the LLaMA-Class Decoder}
\label{sec:beyond-llama-class-decoder}

Decoder-only \transformers remain the reference architecture for most deployed \llms, but they are not the only plausible sequence backbone.\index{state-space model}\index{Mamba}\index{RetNet} Retentive networks, selective state-space models, and hybrid memory architectures try to change the long-context cost model by replacing all-pairs attention with recurrent or chunkwise state updates \cite{sun2023retnet,gu2023mamba,dao2024mamba2,behrouz2025titans}. The attraction is clear: if a model can summarize the prefix into a compact recurrent state, decoding no longer needs a KV cache that grows linearly with every past token in every layer.

These architectures should be compared by contract, not by slogan. Attention exposes direct content-addressed access to previous token representations; recurrent or state-space layers compress history into a state whose update rule determines what can be remembered, forgotten, or retrieved. Mamba-style selective state spaces make the dynamics input-dependent, improving content selectivity relative to older SSMs \cite{gu2023mamba}. Mamba-2 and related state-space duality work further connect attention-like computation and structured recurrent algorithms \cite{dao2024mamba2}. RetNet proposes parallel training with recurrent inference, while memory-augmented designs such as Titans make test-time memory a learned component rather than a fixed KV table \cite{sun2023retnet,behrouz2025titans}.

For a book about \llms, the practical conclusion is conservative. These models are essential frontier material for long-context and streaming systems, but a replacement claim needs evidence at comparable data scale, tokenizer, training budget, context distribution, and serving workload. Many promising sequence models win on asymptotic or medium-scale efficiency but must still prove broad instruction following, retrieval fidelity, tool use, and post-training compatibility at frontier scale.

\section{Evaluation Cautions}
\label{sec:llama-evaluation-cautions}

Architecture claims should be tested under matched training and serving budgets. Replacing LayerNorm with RMSNorm, GELU with SwiGLU, MHA with GQA, or absolute positions with RoPE changes parameter count, activation memory, cache memory, and often the optimal learning rate. A fair ablation controls for tokens, compute, data order, tokenizer, optimizer, and evaluation prompts as much as possible.

Open model reports are valuable primary sources, but they are not interchangeable with independent replication. They often combine architecture changes with data improvements, post-training changes, filtering, safety tuning, and benchmark-specific evaluation choices. When a report attributes gains to an architecture, ask whether the evidence isolates that factor. When a deployment chooses a LLaMA-class model, evaluate the actual checkpoint and decoding path rather than relying on the family name.

\section{Architectural Accounting}
\label{sec:llama-architectural-accounting}

For a dense decoder with $L$ layers, hidden width $d$, attention heads $H_q$, KV heads $H_{\mathrm{kv}}$, head dimension $d_h$, and SwiGLU width $m$, a rough per-layer parameter count is
\begin{equation}
  P_{\mathrm{layer}}
  \approx
  d(H_q d_h) + 2d(H_{\mathrm{kv}}d_h) + d(H_q d_h)
  + 3dm,
\end{equation}
where the first four terms correspond to $W_Q$, $W_K$, $W_V$, and $W_O$, and the last term is the gated feed-forward network. This approximation ignores embeddings, norms, biases, and rounding, but it teaches the key lesson: GQA primarily reduces KV projections and serving cache, whereas the FFN often dominates dense parameters. In an MoE layer, $3dm$ should be replaced by the sum over experts for total parameters and by top-$k$ experts for active parameters. A publishable architecture section should always state which quantity it is reporting.

\section{Key Terms}
\label{sec:chapter-five-key-terms}

\begin{description}[Grouped-query attention]
\item[Activated parameters] Parameters used for a particular token or forward pass, especially in sparse MoE models.
\item[Grouped-query attention] Attention in which multiple query heads share each key-value head.
\item[KV-cache memory] Memory required to store past keys and values for autoregressive decoding.
\item[RoPE] Rotary position embeddings, which rotate query and key dimensions by position-dependent angles.
\item[RMSNorm] Normalization that rescales activations by their root mean square.
\item[SwiGLU] A gated feed-forward layer using a SiLU gate and elementwise multiplication.
\end{description}

\section{Exercises}
\label{sec:chapter-five-exercises}

\begin{enumerate}
\item Modify a GPT-style block on paper into a LLaMA-style block. List every changed component: normalization, bias terms, position handling, MLP activation, and attention cache assumptions.
\item For $d=4096$, compare the approximate MLP parameter counts for a GELU MLP with hidden width $4d$ and a SwiGLU MLP with hidden width $\frac{8}{3}d$ rounded up to the nearest multiple of 256.
\item Derive the relative-position property of RoPE by multiplying $R_m^{\top}R_n$ for a two-dimensional coordinate pair.
\item Compute KV-cache memory for MHA, MQA, and GQA with $L=32$, $B=16$, $T=8192$, $H_q=32$, $H_{\mathrm{kv}}\in\{32,8,1\}$, $d_h=128$, and bfloat16 elements.
\item Design an ablation plan to test whether GQA changes downstream quality independently of total parameter count and training tokens.
\item For an MoE layer with 64 experts, top-2 routing, and capacity limits, name three metrics besides benchmark accuracy that should be reported to make the architecture claim credible.
\end{enumerate}
