\phantomsection%
\addcontentsline{toc}{chapter}{Preface}
\preface

This book is written as a modern, research-grounded introduction to large language models. The prose, organization, notation, examples, and arguments are original. Each chapter is intended to connect three views that are too often taught separately: the mathematical object being optimized, the system that makes the optimization feasible, and the evaluation procedure that determines whether the result is useful.

The book deliberately avoids an implementation-by-implementation narrative. Instead, it follows the lifecycle of a model: data and tokenization, architecture, pretraining, distributed systems, serving, adaptation, alignment, retrieval, reasoning, multimodality, and safety. Historical systems such as GPT, Alpaca, LoRA, RLHF, and LLaMA are treated as conceptual anchors rather than as recipes to copy without context. The intended reader should come away able to ask what is being optimized, what evidence supports the claim, what system constraint changes the result, and what evaluation would make a deployment decision credible.

This is also not a catalog of model names. Model families change quickly, but the contracts behind them are more durable: tokenization defines the interface, pretraining defines the base distribution, post-training defines assistant behavior, inference systems define usable latency and cost, and governance defines what evidence is enough to ship. The chapters therefore emphasize invariants, accounting identities, failure modes, and reproducible records over leaderboard snapshots.

The exposition is evidence-first. Claims about algorithms, scaling behavior, evaluation, or deployment are tied to primary papers or technical reports whenever possible. When a method is still moving, the text separates the stable idea from the particular implementation that made it visible. No third-party slide text, media transcript, dataset record, model output, or code listing is reproduced as manuscript prose. The local course repository and surrounding materials are used as a coverage map and source of exercise ideas, not as text to copy.

Readers can use the book in three passes. A first pass should follow the chapter summaries, key terms, and exercises to build a conceptual map. A second pass should work through the equations, tables, and system diagrams as implementation contracts. A third pass should use the citations, appendix run-card template, and evaluation cautions to design reproducible experiments or review a model report critically.
